Does motif participation provide unique information about species risk of extinction?

### 2021-10-01 manuscript/reorder_main_text.tex
Updates from Overleaf

\documentclass[12pt]{article} 
\usepackage{amsmath} 
\usepackage[dvips]{graphicx}
\usepackage{multirow} 
\usepackage{geometry} 
\usepackage{pdflscape}
\usepackage[labelfont=bf]{caption} 
\usepackage{setspace}
\usepackage[running]{lineno} 
% \usepackage[numbers,sort]{natbib}
\usepackage[round]{natbib} 
\usepackage{array}
\usepackage[table]{xcolor}
\usepackage{xr}

@@

           %\subsection*{Here we test...}
    Specifically, we are interested in whether a consumer's extinction risk depends on 1) the motif profile of the full food web and 2) the motifs in which a focal species participates. 
    In addition, we want to understand how the information provided by motifs is related to classical global and local food web properties in order to better synthesize information provided by different scales of network structure.
    Of the 13 three-species motifs that can occur in empirical food webs, there are four motifs that are over-represented in empirical networks and are thought to contribute to stability \citep{Stouffer2007, Borrelli2015a, giling2019plant}. These four motifs, the three-species chain, omnivory, apparent competition, and direct competition, are also the only ones which can occur in the acyclic networks required for Bayesian network simulations~\citep{Eklof2013}. 
    Therefore, we focus on these four particularly important motifs.

@@

	\subsection*{Calculating persistence}

		First, we assigned all species a baseline probability of extinction ($\pi_{base}$) -- the risk of going extinct given factors not related to the food web itself. 
		For consumer species, this also represents extinction despite all resources being present. 
		Such events can, for example, reflect diseases or stochastic extinctions of small populations.

		As well as this baseline scenario, we simulated a set of scenarios where primary producers had a higher baseline probability of extinction ($\pi_{disturbed}$). 
		These scenarios reflect threats to primary producers such as increased temperatures resulting in droughts, making some species more vulnerable to extinction.
		$\pi_{disturbed}$ ranged between $0.1-0.5$, in steps of $0.08$. 
		The highest disturbance level, $\pi_{disturbed} = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
		Consumer species retained $\pi_{base}=0.1$ in all cases.

		As well as the intrinsic baseline probability of extinction, extinction risk for consumers is affected by access to resources. 
		When all resources are extant, the probability of extinction for a consumer $i$, $P(\lnot i)$, is equal to the baseline probability; $P(\lnot i|X_{1},...,X_{n}) = \pi_{base}$. 
		When all resources are extinct, the consumer will go extinct; $P(\lnot i|\lnot X_{1},...,\lnot X_{n})=1$.

		Calculating $P(i)$ or $P(\lnot i)$ is less straightforward when a consumer $i$ loses a fraction $k/n = f$ of its prey. 
		We used a sigmoid non-linear functional response of consumers to the loss of resources (Appendix \emph{S2}), which has been shown to most accurately capture the secondary extinctions produced by a dynamical model~\citep{Eklof2013}. 
		Probabilities of extinction for resource species are used when calculating probabilities for higher level species.

		Although it is possible to solve the Bayesian network exactly \citep{Eklof2013}, this is cumbersome for larger networks. 
		As numerical simulations are highly efficient and produce the same result \citep{Haussler2020}, we also use simulations here.
		To compute $P(\lnot i|f)$, we calculated the fraction of resources lost ($f$) for each consumer $i$.
		Persistence is determined by comparing $P(\lnot i|f)$ with a randomly drawn number from a uniform distribution between 0-1. 
		If this random number is lower than $P(\lnot i|f)$, the species is considered extinct. 
		The species' marginal probability of persistence is simply $P(i) = 1-P(\lnot i)$.
		To minimize noise due to the random draws, we repeat these calculations 100 000 times per web, with unique random draws when calculating $P(i)$.
		The final persistence for each species is the mean over these runs.

@@

        To provide context for species' motif participation, we also calculated network ``motif profiles'': four-dimensional vectors of the number of each three-species motif in each network~\citep{Stouffer2012}.
        To separate differences in motif structure from differences in network size (larger networks contain more motifs), we normalized motif profiles by dividing the count of each motif by the total count of all motifs. 
        % The motif profile for each network therefore describes the proportions of each motif and has a sum of exactly one.
		We calculated species' motif participation using the Python package \emph{pymfinder}~\citep{pymfinder}.

>>>>

\documentclass[12pt]{article} 
\usepackage{amsmath} 
\usepackage[dvips]{graphicx}
\usepackage{multirow} 
\usepackage{geometry} 
\usepackage{pdflscape}
\usepackage{amsmath}
\usepackage[labelfont=bf]{caption} 
\usepackage{setspace}
\usepackage[running]{lineno} 
% \usepackage[numbers,sort]{natbib}
\usepackage[round]{natbib} 
\usepackage{array}
\usepackage[table]{xcolor}
\usepackage{xr}

@@

           %\subsection*{Here we test...}
    Specifically, we are interested in whether a consumer's extinction risk depends on 1) the motifs in which a focal species participates and 2) the motif profile of the full food web. 
    In addition, we want to understand how the information provided by motifs is related to classical global and local food web properties in order to better synthesize information provided by different scales of network structure.
    Of the 13 three-species motifs that can occur in empirical food webs, there are four motifs that are over-represented in empirical networks and are thought to contribute to stability \citep{Stouffer2007, Borrelli2015a, giling2019plant}. These four motifs, the three-species chain, omnivory, apparent competition, and direct competition, are also the only ones which can occur in the acyclic networks required for Bayesian network simulations~\citep{Eklof2013}. 
    Therefore, we focus on these four particularly important motifs.

@@

	\subsection*{Calculating persistence}	
    % basic idea behind BN modeling
		We use Bayesian Network modeling to calculate the persistence of each species in our simulated networks. 
		A Bayesian Network represent a graphical model for probabilistic relationships among a set of variables. The nodes of the network represents the variables, and the links represent conditional dependencies among the variables. Each node has a set of local probability dependencies, conditional on the state of the parent nodes. 
		In our ecological setting, species act as nodes and is a Bernoulli random variable, being either extant or extinct. The links act as feeding interactions between the nodes.
		The extinction probability of each species is a function of the state of its resources.

		% probability of extinction equation
		Consider a consumer species with several resource species $n$. The extinction probability of this consumer will depend on both the total state of its resource species $n$, and on a baseline extinction probability of extinction, $\pi_{base}$. $\pi_{base}$ represent the risk of going extinct given factors not related to the food web itself, and could reflect diseases or stochastic extinctions of small populations.
		All the species in the network retain a baseline probability of extinction, which can be different to reflect sensitivity of some groups of species.
		If a consumer looses some resources $k$, the fraction of resources lost $f$ will be $f = (k/n)$. Note that we here make a strong assumption, where we only care about the fraction of resources lost and not the identity of the resources species.

		The response of a consumer to the loss of a fraction of its resources can be modelled in several biologically plausible ways. For example, the response could be linear or non-linear.
		A sigmoid non-linear functional response of consumers to the loss of resources ($\alpha >1, \beta >1$) has been shown to most accurately capture the secondary extinctions produced by a dynamical model~\citep{Eklof2013}.
		Therefore, we use a cumulative density function of a beta distribution to get each species \textit{i} probability of extinction:
		\begin{equation}
		\label{betafunc}
		P(\lnot i|f) = \pi_{base} + (1 - \pi_{base}) \frac{B(f;\alpha,\beta)}{B(\alpha,\beta)}
		\end{equation}

		If all resource species of a consumer \textit{C} are present ($f = 0$), the extinction probability will be $P(\lnot C|f) = \pi_{base}$. 
		Similarly, if all resources are extinct ($f = 1$), the extinction probability $P(\lnot C|f)$ will be equal to 1 and the consumer will go extinct.
		If the $f$ is not 1 nor 0, the non-linear sigmoid curve will determine the response of the consumer, which will be more severe if the fraction is high and less severe if the fraction is small. This will in turn affect the extinction probability $P(\lnot C|f)$. 
		For a basal species, we assume that abiotic resources are always available. Therefore, a primary producer is independent of resource species and will always have $f = 0$.

		% how we calculate persistences 
		As consumer species depends on its resource species, we start determining the status of each primary producer in the network.
	    Systematically, we use eqn \ref{betafunc} to receive each species probability of extinction $P(\lnot i|f)$, and next we perform Bernoulli trials. 
	    Imagine a basal species \textit{A} with extinction probability $P(\lnot A) = \pi_{base}$ where $\pi_{base} = 0.1$. We draw a random number $r$ from a uniform distribution between $0-1$. The basal species would be considered extinct if $r < 0.1$, and extant otherwise. 
		We use this information when performing a Bernoulli trial on a consumer \textit{C} connected to this basal species, inserting our information of the $\pi_{base}$ and fraction of resources lost $f$ of this consumer, into eqn \ref{betafunc} to 
	    receive $P(\lnot C|f)$. In the same manner as for the basal species above, we compare $P(\lnot C|f)$ with a random number $r$, and if $r < P(\lnot C|f)$ the consumer species is considered extinct. 
	    In this way, we evaluate each species existence, starting from the bottom and working our way up.
	    As this method is probabilistic, we repeat these calculations 100 000 times per network, with unique random draws.
	    For each species, we then estimate the fraction of simulations in which the species persist, and this value is the approximated probability of persistence.
	    To approximate persistence probabilities through numerical simulation are shown to be highly efficient and produce the same result as exact methods of solving Bayesian networks \citep{Haussler2020}.

    % various disturbance scenarios
        In our baseline scenario, primary producers had a baseline extinction probability of $\pi_{base} = 0.1$. 
		We also simulated a set of scenarios where primary producers had a higher baseline probability of extinction ($\pi_{disturbed}$). 
		These scenarios reflect threats to primary producers such as increased temperatures resulting in droughts, making some species more vulnerable to extinction.
		$\pi_{disturbed}$ ranged between $0.1-0.5$, in steps of $0.08$. 
		The highest disturbance level, $\pi_{disturbed} = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
		Consumer species retained $\pi_{base}=0.1$ in all cases.

		%As well as the intrinsic baseline probability of extinction, extinction risk for consumers is affected by access to resources. 
		%When all resources are extant, the probability of extinction for a consumer $i$, $P(\lnot i)$, is equal to the baseline probability; $P(\lnot i|X_{1},...,X_{n}) = \pi_{base}$. 
		%When all resources are extinct, the consumer will go extinct; $P(\lnot i|\lnot X_{1},...,\lnot X_{n})=1$.

		%Calculating $P(i)$ or $P(\lnot i)$ is less straightforward when a consumer $i$ loses a fraction $k/n = f$ of its prey. 
		%We used a sigmoid non-linear functional response of consumers to the loss of resources (Appendix \emph{S2}), which has been shown to most accurately capture the secondary extinctions produced by a dynamical model~\citep{Eklof2013}. 
		%Probabilities of extinction for resource species are used when calculating probabilities for higher level species.

		%Although it is possible to solve the Bayesian network exactly \citep{Eklof2013}, this is cumbersome for larger networks. 
		%As numerical simulations are highly efficient and produce the same result \citep{Haussler2020}, we also use simulations here.
		%To compute $P(\lnot i|f)$, we calculated the fraction of resources lost ($f$) for each consumer $i$.
		%Persistence is determined by comparing $P(\lnot i|f)$ with a randomly drawn number from a uniform distribution between 0-1. 
		%If this random number is lower than $P(\lnot i|f)$, the species is considered extinct. 
		%The species' marginal probability of persistence is simply $P(i) = 1-P(\lnot i)$.
		%To minimize noise due to the random draws, we repeat these calculations 100 000 times per web, with unique random draws when calculating $P(i)$.
		%The final persistence for each species is the mean over these runs.

@@

        To provide context for species' motif participation, we also calculated network ``motif profiles'': four-dimensional vectors of the number of each three-species motif in each network~\citep{Stouffer2012}.
        To separate differences in motif structure from differences in network size (larger networks contain more motifs), we normalized motif profiles by dividing the count of each motif by the total count of all motifs. 
		We calculated species' motif participation using the Python package \emph{pymfinder}~\citep{pymfinder}.
